Two Decades of Smalltalk VM Development

### 2018-08-17 paper.tex
Added section on Arbitrary breakpoints, history, and decorated disassembly

In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of native code it generates. The JIT itself is written in Slang and simulated with the Smalltalk execution model. To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6) were implemented so that the machine code can be simulated. Calls in-between slang code and direct machine code are a little bit trickier to simulate. Calls from machine code to slang code are implemented by using multiple invalid processor instructions, leading to a trap in the processor simulator. Such traps can be caught is the VM simulator, which then resumes Slang simulation by calling the correct method based on the invalid processor instruction. Calling machine code from Slang requires to start the machine code simulation but also to raise an exception to stop Slang simulation. Indeed, in the production VM, the processor can execute the C code or the native code generated by the JIT, but not both at the same time.

Machine code simulation can be performed in two different ways. The processor simulator can start simulating code until it meets an invalid instruction. That version is convenient because it is the fastest to execute, while still catching errors such as invalid memory accesses. Alternatively, it can simulate one instruction at a time. This second version is slower, but it allows to implement specific debugging features, such as conditional breakpoints in-between each machine instruction.

@@

The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects.  As code is JITted and specific variables and runtime routines are referenced the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction \footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it will attempt to raise an exception.  Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  The processor simulator is invoked in the context of an exception handler for ProcessorException exceptions which invokes a Smalltalk method, handleSimulationTrap:, that invokes the relevant closure.  Once the closure has completed, the processor simulator is advanced to the nextpc and simulation resumes.  In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM.

\paragraph{Machine code decoration, arbitrary breakpoints and history}
Blah, blah, blah.

@@

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes around 15 seconds on a recent Macbook pro. While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid.

To work around this problem, we implemented a tool called \emph{In-image compilation}. In-image compilation basically allows to call the JIT as a Smalltalk library on a given bytecode compiled method to generate the corresponding machine code, calls the bound processor simulator to disassemble the code and decorate the disassembly with Smalltalk specific information. To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To work around this, we built a facade, which includes mock addresses for all the object the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.

@@

@@

The simulator has several limitations. Due to the simulation infrastructure being different from the actual hardware, code is run differently and one could think it leads to strange bugs happening only in simulation or in production. However, we have been using this infrastructure since 1995, and this kind of bugs are really rare and usually easy to fix. We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot be simulated.

>>>>

In addition to the interpreter simulator, simulating the JIT requires simulate the execution of the native code it generates. The JIT itself is written in Slang and simulated with the Smalltalk execution model. To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, SMalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely. All but the MIPSLE simulators are accessed through SMalltalk primitives.  These are effectively a combination of an FFI call to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. The next two sections describe in detail how this scheme is used to ensure that simulation and production machine code agree, how simulation machien code is interfaced to the rest of teh simulation, and the debugging facilities for machine code that are available.

%Calls in-between slang code and direct machine code are a little bit trickier to simulate. Calls from machine code to slang code are implemented by using multiple invalid processor instructions, leading to a trap in the processor simulator. Such traps can be caught is the VM simulator, which then resumes Slang simulation by calling the correct method based on the invalid processor instruction. Calling machine code from Slang requires to start the machine code simulation but also to raise an exception to stop Slang simulation. Indeed, in the production VM, the processor can execute the C code or the native code generated by the JIT, but not both at the same time.

%Machine code simulation can be performed in two different ways. The processor simulator can start simulating code until it meets an invalid instruction, or until the VM's internal heartbeat ticks. That version is convenient because it is the fastest to execute, while still catching errors such as invalid memory accesses. Alternatively, it can simulate one instruction at a time. This second version is slower, but it allows to implement specific debugging features, such as conditional breakpoints in-between each machine instruction.

@@

The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects.  As code is JITted and specific variables and runtime routines are referenced the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction \footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it will attempt to raise an exception.  Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  The processor simulator is invoked in the context of an exception handler for ProcessorException exceptions which invokes a Smalltalk method, handleSimulationTrap:, that invokes the relevant closure.  Once the closure has completed, the processor simulator is advanced to the nextpc and simulation resumes.  In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM. Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation portable and processor independent.

\paragraph{Arbitrary breakpoints, history, and decorated disassembly}
The JIT maintains a simulation-only variable that determines whether the procerssor smulator is invoked in single-stepping mode, or run-until-exception-or-interrupt mode.  When in single stepping mode a circular buffer remembers N previous instructions, allowing one to examine an arbitrary number of instructions (by default 160) preceeding some error.  The arbitrary breakpoint facility is intelligent enough to only enable single stepping once code has been JITted at that address, using run mode until that point.

There is a small polymorphic scheme for breakpoints.  Booleans, integers, and arrays understand isBreakpointFor: pc.  So if the breakPC object is false there is no breakpoint.  If the breakPC is true then any pc is a potential breakpoint.  If it is an integer then that pc is a potential breakpoint.  If the pc is an Array then the pc is a breakpoint if any value in the array answers true to isBreakpointFor:.  The scheme could trivially be extended to include intervals.  If breakPC isBreakpointFor: pc is true then the breakBlock is sent shouldStopIfAtPC: pc.  booleans and closures understand shouldStopIfAtPC:.  true and false make the breakpoint unconditional or disabled respectively, and a closure evaluates itself, allowing one to specify arbitrarily complex breakpoints.  For example, one can specify that execution should stop at a particular pc if the top two elements on the stack satisfy some criterion.  It is much more powerful and much simpler to use than typical machine-level debuggers, and it can be extended as one is debugging (for example, adding interval support during the middle of a debugging session if required).

Each processor simulator is expected to be able to disassemble its machine code.  The Smalltalk code for each processor simulator implements a simple parser that identifies constants and field offsets embedded in machine code and passes these to routines in the JIT and the interpreter that lookup addresses, matching them to objects, in particular message selectors and classes, and to method local variable names, rendering the machine code much more readable than machine code displayed in a typical machine oriented debugger.  For example, here is a snippet of decorated disassembly for x64:

\lstdefinelanguage
   [x64]{Assembler}     % add a "x64" dialect of Assembler
   [x86masm]{Assembler} % based on the "x86masm" dialect
   % with these extra keywords:
   {morekeywords={movq, cmpq, %
                  rax,rdx,rcx,rbx,rsi,rdi,rsp,rbp, %
                  r8,r8d,r8w,r8b,r9,r9d,r9w,r9b, %
                  r10,r10d,r10w,r10b,r11,r11d,r11w,r11b, %
                  r12,r12d,r12w,r12b,r13,r13d,r13w,r13b, %
                  r14,r14d,r14w,r14b,r15,r15d,r15w,r15b}} % etc.

\lstset{language=[x64]Assembler}

\begin{lstlisting}
00002063: movq 0x800(%rbx), %rax = 'stackLimit' : 48 8B 83 00 08 00 00 
0000206a: cmpq %rax, %rsp : 48 39 C4 
0000206d: jb .-0x47 (0x2028=indexOf:startingAt:ifAbsent:@28) : 72 B9 
HasBytecodePC bc 40/41:
0000206f: movq start@20(%rbp), %rsi : 48 8B 75 14 
00002073: movq anElement@28(%rbp), %rdi : 48 8B 7D 1C 
00002077: movq self@-12(%rbp), %rdx : 48 8B 55 F4 
0000207b: movq $0x0, %rcx : 48 C7 C1 00 00 00 00 
00002082: call .-0x1B2F (0x558=ceSend2Args) : E8 D1 E4 FF FF 
IsSendCall indexOf:startingAt: bc 44/45:
00002087: movq %rdx, index@-20(%rbp) : 48 89 55 EC 
\end{lstlisting}

This code is the end of the entry sequence that checks the stack pointer against the stack limit for the end of the current stack page, which is also used to break out of machine code to service interrupts or events such as invoking the scavenger, etc.  \emph{\%rbx} is used as a base register pointing at the variables in interpreter.  Following that is the code for \emph{index := self indexOf: anElement startingAt: start} where \emph{anElement} and \emph{start} are method parameters and \emph{index} is a local variable.  HasBytecodePC and IsSendCall are decorations that identify important locations in machine code, including object references, runtime calls, etc. These points are specified using metadata implemented as a simple bytecoded language, a stream of such bytes being added to the end of each machine code method.  This metadata is parsed when the garbage collector needs to locate object references in machine code, when linked sends must be relocated when the machine code zone is compacted to make space by throwing away some number of least recently used methods, and when machine code pcs must be mapped to their corresponding bytecode pcs to preserve the illusion that the VM is a bytecode interpreter using a spaghetti stack of Context objects.

@@

Simulating the whole VM requires going through the whole start-up sequence: loading the snapshot, running code registered in the start-up sequence and resuming the user interface. The whole start-up takes around 15 seconds on a recent Macbook pro. While developing the JIT, this start-up time may still be too long and move the live programming experience to an edit-compile-run cycle, which we want to avoid.  Further, a particular method that exercises some new part of the JIT may be problematic to have executed suring simulation and hence not easily JITted during simulation.

To work around these problems, we implemented a tool called \emph{In-image compilation}. In-image compilation allows invoking the JIT as a Smalltalk library on any bytecode compiled method in the host Smalltalk system (i.e. \emph{not} in the simulated heap) to generate the corresponding machine code. In-image compilation invokes the JIT on the method, calls the bound processor simulator to disassemble the code and decorate the disassembly and then dumps the output in a text wwindow. Indeed the above snippet is from an in-image compilation.  To generate the machine code, the JIT has to access specific objects (the compiled method, the literals, known objects such as true, false or nil) as if they were in the simulated memory. To implement this we built a facade, which masquerades as the simulation's memory manager, and translates the state of the bytecode compiled method in the current simulation into state that makes it appear as if it were resident in a simulaiton byte array.  This includes mock addresses for all the object the JIT may require to generate the machine code of a given method. Figure \ref{fig:InImageCompiler} summarizes the process. Note that this technique applies for the baseline JIT, which translate a single bytecode method into machine code, adaptive optimizations and speculative optimizations are debugged differently.

@@

Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecoded method from the assembly code.  Although for methods of the complexity produced by the adaptive optimiser this is hardly pleasurable.

@@

The simulator has several limitations. Due to the simulation infrastructure being different from the actual hardware, code is run differently and one could think it leads to strange bugs happening only in simulation or in production. However, we have been using this infrastructure since 1995, and these kinds of bugs are really rare and usually easy to fix. We have however two main limitations: simulation performance which is quite slow and calls to external C/C++/machine code which cannot (yet) be simulated.
